Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-01 analysis/paper/paper.Rmd
near final methods edits

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (RKM) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event and upslope instability that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973]. That sediment formed an alluvial fan which created a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the juvenile DSR and NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Northern Pikeminnow were surveyed near the peaks of the fall DSR and spring NRR emigrations. The initial intent was to survey during the emigration of the 2018 brood year (fall 2019 and spring 2020), however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until fall 2020. Although this allowed an additional survey during the fall DSR emigration, surveys during the spring NRR emigration were still lacking. Consequently, an additional week of surveying was added in spring 2021. Fall surveys occurred over two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were abandoned in subsequent years. <!-- For example, fish were able to elude the electrical field of the raft electrofisher due to 1) the lack of flow in Deadwater Slough and 2) our inability to "chase" fish downstream as they evaded the electricity. Snorkeling was ineffective due to the high turbidity of the mainstem Salmon River and Deadwater Slough area. Significant amounts of debris in the water column also created concerns that fyke nets could become entangled. Additionally, there were concerns that raft electrofishers and fyke nets could incidentally capture ESA-listed adult steelhead that are present in the mainstem Salmon River during fall and spring. -->After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present in the Salmon River during our surveys.<!--, but recognizing that angling can have both species and size selectivity (bias).--> Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. Each survey day, anglers boated or hiked their catch (periodically or upon filling a livewell) to a processing station at the boat ramp, approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and then were released. Released fish were distributed throughout Deadwater Slough to facilitate mixing back into the population. Fish that died prior to release were included in the total capture (or recapture) count, but not included in the number of marked fish available for recapture after that date. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to calculate CPUE as number of Northern Pikeminnow caught per angler hour.

@@

Two types of mark-recapture estimators were used to explore variation in abundance estimates: a single census and a multiple census estimator. Estimators fell into two broad categories: single census and multiple census. For the single census estimator, we treated the first week of the survey as the mark event and the following week as the recapture event, pooling data within each event week. Alternatively, the multiple census estimator treated each day as an event and use information about the total number of marked fish from all previous events to calculate the total abundance.

@@

<!--
$$
\hat{N} = \frac{(M)(n)}{(m)}
$$
-->
$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999]. The Schnabel estimator is essentially a weighted average of a series of Chapman-modified Lincoln-Peterson estimators.

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those locales, marked fish may have been less susceptible to capture. To account for this potential bias, we adjusted the Schnabel estimator with an assumption that marked and released fish during each survey day were not available for recapture for 48 hours.

@@

<!-- delayed mixing Schnabel equation here; include statement about fish being "hook shy" -->

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and 2020 surveys. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall survey to the total CPUE in the spring survey, then multiplied that ratio by the average abundance from the fall surveys.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into a linear density by dividing by the length of Deadwater Slough (km) measured in GIS software and using drone-generated orthography.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks  to be analyzed in a controlled environment. For each stomach sample, wet weight (grams) was recorded for the total stomach contents, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones, or were categorized as unknown [@Hansel1988; @Frost2000]. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology.

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because were were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average length of Northern Pikeminnow caught in Deadwater Slough during our study was converted to weight using a weight-length formula from @Parker1995.

@@

This resulted in average spring and fall starting eights of 504.4 and 670.5 g for Northern Pikeminnow with average lengths of `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 RKM downstream of Deadwater Slough.

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted separate model runs that assume different proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey. 14 models were run (7 diet scenarios and 2 seasons), each providing an estimate of the total grams of fish consumed by an individual Northern Pikeminnow (***table reference***). To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population size during fall and spring.

@@

The proportion of Chinook Salmon prey relative to other fish prey in the Northern Pikeminnow diet is also unknown. However, there is some evidence that juvenile salmonids are by far the most consumed fish prey [@Zimmerman1999; @Shively1996]. Moreover, it is estimated that Chinook Salmon make up 64.2%, 29.3%, and 49.3% of the fish prey consumed by Northern Pikeminnow in the Columbia River below Bonneville Dam, Columbia River reservoirs, and in the lower Snake River, respectively [@Zimmerman1999]. Because the bioenergetics model does not differentiate juvenile Chinook Salmn from other fish prey items, we performed a sensitivity analysis to assess potential impacts of Northern Pikeminnow predation on local Chinook Salmon populations. Using values similar to @Zimmerman1999 we modeled three diet scenarios where Chinook Salmon comprised 30%, 50%, and 65% of the fish prey, by weight, in the Northern Pikeminnow diet during the fall DSR and spring NRR emigration periods. In this analysis, we assume 60% of the Northern Pikeminnow's total diet is comprised of fish, representing the median in our range of modeled diets.

@@

Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is often adult returns. We assessed the impacts on adult returns by estimating the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. This calculation was made by multiplying the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

>>>>

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event and upslope instability that deposited substantial amounts of sediment at the confluence of the Salmon River [@Emerson1973]. That sediment formed an alluvial fan which created a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were by far the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. The initial intent was to survey during the emigration of the 2018 brood year (fall 2019 and spring 2020), however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until fall 2020. Although this allowed an additional survey during the fall DSR emigration, surveys during the spring NRR emigration were still lacking. Consequently, an additional week of surveying was added in spring 2021. The fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration/emigration.

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were abandoned in subsequent years. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present in the Salmon River during our surveys. Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted.

Our study relied heavily on volunteer anglers who were permitted to fish anywhere within Deadwater Slough. Each survey day, anglers boated or hiked their catch (periodically or upon filling a livewell) to a processing station at the boat ramp, approximately 500 m downstream from the top of the slough. For each fish, we recorded the date of capture, species, total length (TL; mm), and whether the fish was previously marked. Unmarked fish were given a physical mark (e.g., hole punch of lower caudal, upper caudal, left pelvic, right pelvic) unique to each day and then were released. Released fish were distributed throughout Deadwater Slough to facilitate mixing back into the population. Fish that died prior to release were included in the total capture (or recapture) count, but not included in the number of marked fish available for recapture after that date. Finally, we recorded the angling start and end time for each crew (person or combinations of persons) to calculate CPUE as the number of Northern Pikeminnow caught per angler hour.

@@

Two types of mark-recapture estimators were used to explore variation in abundance estimates: a single census and a multiple census estimator. For the single census estimator, we treated the first week of the survey as the mark event and the following week as the recapture event, pooling data within each week. Alternatively, the multiple census estimator treated each day as an event and used information about the total number of marked fish from all previous events to calculate the total abundance.

@@

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator, but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999]. The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those preferred locales, marked fish may have been less susceptible to capture than unmarked fish. To account for this potential bias, we also used an adjusted the Schnabel estimator which assumed that marked and released fish during each survey day were not available for recapture for 48 hours. The adjusted Schnabel estimator, which we refer to as the "delayed mixing" Schnabel estimator, would also account for marked fish being "hook shy" for 48 hours following release.

@@

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and 2020 surveys. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall survey to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into a linear density by dividing by the length of Deadwater Slough (rkm) measured in GIS software and using drone-generated orthography.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks  to be analyzed in a controlled environment. For each stomach sample, wet weight (grams) was recorded for the total stomach contents, including all non-fish items (e.g., macroinvertebrates, organic matter), as well as the portion consisting of fish. Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology.

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average length of Northern Pikeminnow caught in Deadwater Slough during our study was converted to weight using a weight-length formula from @Parker1995.

@@

This resulted in average spring and fall starting weights of 504.4 and 670.5 g for Northern Pikeminnow with average lengths of `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 rkm downstream of Deadwater Slough.

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted separate model runs that assume different proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey. Fourteen model runs were completed (7 diet scenarios and 2 seasons), each providing an estimate of the total grams of fish consumed by an individual Northern Pikeminnow (***table reference***). To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall and spring.

@@

The proportion of Chinook Salmon prey relative to other fish prey in the Northern Pikeminnow diet is also unknown. However, there is some evidence that juvenile salmonids are by far the most consumed fish prey [@Zimmerman1999; @Shively1996]. Moreover, it is estimated that Chinook Salmon make up 64.2%, 29.3%, and 49.3% of the fish prey consumed by Northern Pikeminnow in the Columbia River below Bonneville Dam, in Columbia River reservoirs, and in the lower Snake River, respectively [@Zimmerman1999]. Because the bioenergetics model does not differentiate juvenile Chinook Salmon from other fish prey items, we performed a sensitivity analysis to assess potential impacts of Northern Pikeminnow predation on local Chinook Salmon populations. Using values similar to @Zimmerman1999, we modeled three diet scenarios where Chinook Salmon comprised 30%, 50%, and 65% of the fish prey, by weight, in the Northern Pikeminnow diet during the fall DSR and spring NRR emigration periods. In this analysis, we assume 60% of the Northern Pikeminnow's total diet is comprised of fish, representing the median scenario in our range of modeled diets.

@@

Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is often adult returns. We assessed the potential impacts on adult returns by estimating the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. This calculation was made by multiplying the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.
